\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-03-24 k-planar.tex
minor update
minor edits and comments

\noindent\cref{PlanarProduct} has been the key tool in solving the following well-known open problems:
\begin{compactitem}
\item \citet{DJMMUW20} use it to prove that planar graphs have bounded queue-number (resolving a conjecture of \citet{HLR92}).
\item  \citet{dujmovic.esperet.ea:planar} use it to prove that planar graphs have bounded non-repetitive chromatic number (resolving a conjecture of \citet{AGHR-RSA02}).
\item \citet{DFMS21} use it to prove that planar graphs have $p$-centered chromatic number $O(p^2\log p)$ and give a matching lower bound.
\item \citet{DEJGMM21} use it to find asymptotically optimal adjacency labellings of planar graphs (resolving a problem of \citet{kannan.naor.ea:implicit}).
\item \citet{EJM} use it to show the existence of a 'universal graph' with $n^{1+o(1)}$ vertices and edges that contains every $n$-vertex planar graph as an induced subgraph (resolving a problem of \citet{babai.chung.ea:on}).
\end{compactitem}
In addition, \cref{PlanarProduct} has been used to resolve or make substantial progress on a number of other problems on planar graphs, including $\ell$-vertex ranking~\citep{BDJM}, twin-width~\citep{BKW}, and odd colourings~\citep{DMO}.

@@

For at least two applications, the planarity of $H$ is critical in obtaining asymptotically tight bounds \cite{BDJM,DFMS21}. \cref{1_planar_product} is proven in the more general setting of $d$-framed multigraphs. This more general result implies a product structure theorem for $d$-map graphs in which $H$ is also planar.

@@

\cref{summary_table} summarizes existing results on product structure theorems for minor-closed graph classes and new results for non-minor-closed graph classes.  The remainder of this paper is organized as follows:

\begin{table}
  \begin{center}
    \begin{tabular}{lccl}
      \textbf{Graph class} & \boldmath $\tw(H)$ & \boldmath $\ell$ & \textbf{Ref.} \\ \hline\hline
      Planar & $3^*$ & $3$ & \cite{DJMMUW20} \\
      Planar & $6^*$ & $1$ & \cite{UWY} \\
      Genus-$g$ & $3^*$ & $\max\{2g,3\}$ & \cite{DHHW} \\
      Genus-$g$ & $2g+6$ & 1 & \cite{UWY} \\
      Apex-minor-free & $O(1)$ & $O(1)$ & \cite{DJMMUW20} \\[1ex]
      % B-minor-free & $O(1)$ & $O(1)$ & \cite{DJMMUW20} \\
      $k$-planar & $\binom{k+4}{3}-1$ & $18k^2 + 48k + 30$ & \cref{kPlanarProduct} \\
      $(g,k)$-planar & $\binom{k+4}{3}-1$ & $\max\{2g,3\}(6k^2+16k+10)$ & \cref{gkPlanarProduct} \\
      $(g,\delta)$-string & $\binom{\delta+4}{3}-1$ & $\max\{2g,3\}(\delta^4 + 4\delta^3 + 9\delta^2 + 10\delta+4)$ & \cref{StringProduct} \\
      $k$-NN & $O(k^6)$ & $O(k^4)$ & \cref{nn_product_structure} \\
      $d$-framed & $3^*$ & $d+ 3\floor{d/2} - 3$ & \cref{d_framed_product_stucture} \\
      $1$-planar & $3^*$ & $7$ & \cref{1_planar_product} \\
      $d$-map & $3^*$ & $d+ 3\floor{d/2} - 3$ & \cref{dMapProduct} \\
      $(g,d)$-map & $9$ & $\max\{2g,3\}\,(7d^2 -21d)$ & \cref{gdMapProduct} \\ \hline
      \\[.2ex]
      \multicolumn{4}{l}{$^*$ the bound also applies to the simple treewidth of $H$.}
    \end{tabular}
  \end{center}
  \caption{Known product structure theorems (of the form $G\subseteq H\boxtimes P\boxtimes K_\ell$).}
  \label{summary_table}
\end{table}

\begin{compactitem}
  \item In \cref{Structure} we prove our main result for shortcut systems (\cref{ShortcutProduct}), and its optimisations for $k$-planar graphs and their genus-$g$ generalisation (\cref{kPlanarProduct,gkPlanarProduct}).
  \item \Cref{examples} uses \cref{ShortcutProduct,kPlanarProduct,gkPlanarProduct} to derive product structure theorems for string graphs, powers of bounded-degree graphs, map graphs, and $k$-nearest neighbour graphs.
  \item \Cref{FramedSection} proves a product structure theorem for $d$-framed multigraphs in which the graph $H$ has treewidth $3$ and is planar.  Using this result, we derive improved product structure theorems for $1$-planar graphs and planar map graphs, again with a planar graph $H$.
  \item \Cref{Applications} quickly surveys applications of product structure and the consequences of the current work.
  \item \cref{WrappingUp} concludes with an open problem.
\end{compactitem}

@@

The definition of $k$-planar graphs naturally generalises to genus-$g$ surfaces. A graph $G$ embedded on a surface $\Sigma$ is $(\Sigma,k)$-plane if every edge of $G$ is involved in at most $k$ crossings.  A graph $G$ is $(g,k)$-planar if it is isomorphic to some $(\Sigma,k)$-plane graph, for some surface $\Sigma$ with Euler genus at most $g$. \cref{AddDummy} immediately generalises as follows:

@@

\cref{d_framed_product_stucture} follows quickly from the following technical lemma, which is an extension of the analogous result for plane graphs \cite{DJMMUW20}. The following definition is a convenient way to establish the planarity of $H$ in \cref{d_framed_product_stucture}:  A tree-decomposition $\mathcal{T}$ of a graph $G$ is \defin{$t$-simple} if each of its bags has size at most $t+1$ and, for each $t$-element subset $\{v_1,\ldots,v_t\}\subseteq V(G)$, at most two bags of $\mathcal{T}$ contain $\{v_1,\ldots,v_t\}$.  Finally, a \defin{BFS spanning forest} $T$ of $G$ rooted at a set $V_0\subseteq V(G)$ is a spanning forest of $G$ with the property that for each $v\in V(G)$,  \[\min\{\dist_G(v,w):w\in V_0\}=\min\{\dist_T(v,w):w\in V_0\}.\]

@@

\begin{cor}
\label{dMapQueue}
Every $d$-map graph has queue-number at most 
$15 (d+3\floor{d/2}-3)  + \floor{\tfrac{3}{2}(d+3\floor{d/2}-3)}$.
\note{DW}{I have updated this bound using the better qn bound for planar 3-trees. Note that with $d=3$ we get the original bound of 49.}
\end{cor}

@@

Combining \cref{non-repetitive} with our results on product structure, we obtain the following corollary:
\begin{cor}\label{nonrepetitive_cor}
  The following bounds hold for the nonrepetitive chromatic number of any graph from each of the following classes:
  \begin{compactenum}
    \item $k$-planar: $(18k^2+48k+30)\cdot 4^{\binom{k+4}{3}}$
    \item $(g,k)$-planar: $\max\{2g,3\}\cdot(6k^2+16k+10) 4^{\binom{k+4}{3}}$
    % \item $(g,\delta)$-string: $g4^{O(\delta^3)}$
    \item $1$-planar: $1792$
    \item $d$-map: $256(d+3\floor{d/2}-3)$
    % \item $(g,d)$-map: $g4^{O(d)}$
    % \item $k$-NN: $4^{O(k^6)}$
    \item $(g,d)$-map graph: $(7d^2-21d)\cdot 4^{10}$
    % \item
  \end{compactenum}
\end{cor}

@@

For planar graphs of treewidth at most $3$ an $O(p^2\log p)$ upper bound is known:

@@

\begin{cor}\label{p_centered_cor}
  The following bounds hold for the $p$-centered chromatic number of any graph from each of the following classes:
  \begin{compactenum}
    \item $k$-planar: $(18k^2+48k+30)\cdot 4^{\binom{k+4}{3}}$
    \item $(g,k)$-planar: $\max\{2g,3\}\cdot(6k^2+16k+10)(p+1)\binom{p+\binom{k+4}{3}-1}{\binom{k+4}{3}-1}$
    % \item $(g,\delta)$-string: $g4^{O(\delta^3)}$
    \item $1$-planar: $7(p+1)^2(p\ceil{\log(p+1)}+2p+1)$
    \item $d$-map: $(d+3\floor{d/2}-3)\cdot (p+1)^2(p\ceil{\log(p+1)}+2p+1)$
    % \item $(g,d)$-map: $g4^{O(d)}$
    % \item $k$-NN: $4^{O(k^6)}$
    \item $(g,d)$-map: $\max\{2g,3\}\cdot 7d(d-3)(p+1)\cdot \binom{p+9}{9}$
    % \item
  \end{compactenum}
\end{cor}

@@

>>>>

\noindent\cref{PlanarProduct} has been the key tool in solving the following well-known open problems:
\begin{compactitem}
\item \citet{DJMMUW20} use it to prove that planar graphs have bounded queue-number (resolving a conjecture of \citet{HLR92}).
\item  \citet{dujmovic.esperet.ea:planar} use it to prove that planar graphs have bounded non-repetitive chromatic number (resolving a conjecture of \citet{AGHR-RSA02}).
\item \citet{DFMS21} use it to prove that planar graphs have $p$-centered chromatic number $O(p^2\log p)$ and give a matching lower bound.
\item \citet{DEJGMM21} use it to find asymptotically optimal adjacency labellings of planar graphs (resolving a problem of \citet{kannan.naor.ea:implicit}).
\item \citet{EJM} use it to show the existence of a `universal graph' with $n^{1+o(1)}$ vertices and edges that contains every $n$-vertex planar graph as an induced subgraph (resolving a problem of \citet{babai.chung.ea:on}).
\end{compactitem}
In addition, \cref{PlanarProduct} has been used to resolve or make substantial progress on a number of other problems on planar graphs, including $\ell$-vertex ranking~\citep{BDJM}, twin-width~\citep{BKW}, and odd colourings~\citep{DMO}.

@@

For at least two applications, the planarity of $H$ in \cref{1_planar_product} is critical in obtaining asymptotically tight bounds \cite{BDJM,DFMS21}. \cref{1_planar_product} is proven in the more general setting of $d$-framed multigraphs. This more general result also implies a product structure theorem for $d$-map graphs in which $H$ is a planar.

@@

\cref{summary_table} summarizes existing results on product structure theorems for minor-closed graph classes and new results for non-minor-closed graph classes.

\begin{table}[h]
  \begin{center}
    \begin{tabular}{lccl}
      \textbf{Graph class} & \boldmath $\tw(H)$ & \boldmath $\ell$ & \textbf{Reference} \\ \hline
      planar & $3^*$ & $3$ & \cite{DJMMUW20} \\
      planar & $6^*$ & $1$ & \cite{UWY} \\
      genus $g$ & $3^*$ & $\max\{2g,3\}$ & \cite{DHHW} \\
%      genus $g$ & $2g+6$ & 1 & \cite{UWY} \\
      apex-minor-free & $O(1)$ & $O(1)$ & \cite{DJMMUW20} \\[1ex]
      % B-minor-free & $O(1)$ & $O(1)$ & \cite{DJMMUW20} \\
      $k$-planar & $\binom{k+4}{3}-1$ & $18k^2 + 48k + 30$ & \cref{kPlanarProduct} \\
      $(g,k)$-planar & $\binom{k+4}{3}-1$ & $\max\{2g,3\}(6k^2+16k+10)$ & \cref{gkPlanarProduct} \\
      $(g,\delta)$-string & $\binom{\delta+4}{3}-1$ & $\max\{2g,3\}(\delta^4 + 4\delta^3 + 9\delta^2 + 10\delta+4)$ & \cref{StringProduct} \\
      $k$-NN & $O(k^6)$ & $O(k^4)$ & \cref{nn_product_structure} \\
      $d$-framed & $3^*$ & $d+ 3\floor{d/2} - 3$ & \cref{d_framed_product_stucture} \\
      $1$-planar & $3^*$ & $7$ & \cref{1_planar_product} \\
      $d$-map & $3^*$ & $d+ 3\floor{d/2} - 3$ & \cref{dMapProduct} \\
      $(g,d)$-map & $9$ & $\max\{2g,3\}\,(7d^2 -21d)$ & \cref{gdMapProduct} \\ \hline
      \\[.2ex]
      \multicolumn{4}{l}{$^*$ the bound also applies to the simple treewidth of $H$.}
    \end{tabular}
  \end{center}
  \caption{Known product structure theorems (of the form $G\subseteq H\boxtimes P\boxtimes K_\ell$).}
  \label{summary_table}
\end{table}

The remainder of this paper is organized as follows:
\begin{compactitem}
  \item In \cref{Structure} we prove our main result for shortcut systems (\cref{ShortcutProduct}), and its optimisations for $k$-planar graphs and their genus-$g$ generalisation (\cref{kPlanarProduct,gkPlanarProduct}).
  \item \Cref{examples} uses \cref{ShortcutProduct,kPlanarProduct,gkPlanarProduct} to derive product structure theorems for string graphs, powers of bounded-degree graphs, map graphs, and $k$-nearest neighbour graphs.
  \item \Cref{FramedSection} proves a product structure theorem for $d$-framed multigraphs in which the graph $H$ has treewidth $3$ and is planar.  Using this result, we derive improved product structure theorems for $1$-planar graphs and planar map graphs, again with a planar graph $H$.
  \item \Cref{Applications} quickly surveys applications of product structure and the consequences of the current work.
  \item \cref{WrappingUp} concludes with an open problem.
\end{compactitem}

@@

The definition of $k$-planar graphs naturally generalises to genus-$g$ surfaces. A graph $G$ embedded on a surface $\Sigma$ is \defin{$(\Sigma,k)$-plane} if every edge of $G$ is involved in at most $k$ crossings.  A graph $G$ is \defin{$(g,k)$-planar} if it is isomorphic to some $(\Sigma,k)$-plane graph, for some surface $\Sigma$ with Euler genus at most $g$. \cref{AddDummy} immediately generalises as follows:

@@

\cref{d_framed_product_stucture} is a consequence of the following technical lemma, which is an extension of the analogous result for plane graphs~\cite{DJMMUW20}. The following definition is a convenient way to establish the planarity of $H$ in \cref{d_framed_product_stucture}:  A tree-decomposition $\mathcal{T}$ of a graph $G$ is \defin{$t$-simple} if each of its bags has size at most $t+1$ and, for each $t$-element subset $\{v_1,\ldots,v_t\}\subseteq V(G)$, at most two bags of $\mathcal{T}$ contain $\{v_1,\ldots,v_t\}$.  Finally, a \defin{BFS spanning forest} $T$ of $G$ rooted at a set $V_0\subseteq V(G)$ is a spanning forest of $G$ with the property that for each $v\in V(G)$,  \[\min\{\dist_G(v,w):w\in V_0\}=\min\{\dist_T(v,w):w\in V_0\}.\]

@@

\begin{cor}
\label{dMapQueue}
Every $d$-map graph has queue-number at most 
$\floor{ \frac{33}{2} (d+3\floor{d/2}-3) }$.
\note{DW}{I have updated this bound using the better qn bound for planar 3-trees. Note that with $d=3$ we get the original bound of 49.}
\end{cor}

@@

Combining \cref{non-repetitive} with our results on product structure, we obtain the following corollary:
\begin{cor}\label{nonrepetitive_cor}
  The following bounds hold for the nonrepetitive chromatic number of any graph from each of the following classes:
  \begin{compactenum}
    \item $k$-planar: $(18k^2+48k+30)\cdot 4^{\binom{k+4}{3}}$
    \item $(g,k)$-planar: $\max\{2g,3\}\cdot(6k^2+16k+10) 4^{\binom{k+4}{3}}$
    % \item $(g,\delta)$-string: $g4^{O(\delta^3)}$
    \item $1$-planar: $1792$
    \item $d$-map: $256(d+3\floor{d/2}-3)$
    % \item $(g,d)$-map: $g4^{O(d)}$
    % \item $k$-NN: $4^{O(k^6)}$
    \item $(g,d)$-map graph: $(7d^2-21d)\cdot 4^{10}$
  \end{compactenum}
\note{DW}{I prefer bullets here}
\end{cor}

@@

For planar graphs of treewidth at most $3$ an $O(p^2\log p)$ upper bound is known:
\note{DW}{I prefer ``a $O(p^2\log p)$'', pronounced ``a big-O...''}

@@

\begin{cor}\label{p_centered_cor}
  The following bounds hold for the $p$-centered chromatic number of any graph from each of the following classes:
  \begin{compactenum}
    \item $k$-planar: $(18k^2+48k+30)\cdot 4^{\binom{k+4}{3}}$
    \item $(g,k)$-planar: $\max\{2g,3\}\cdot(6k^2+16k+10)(p+1)\binom{p+\binom{k+4}{3}-1}{\binom{k+4}{3}-1}$
    % \item $(g,\delta)$-string: $g4^{O(\delta^3)}$
    \item $1$-planar: $7(p+1)^2(p\ceil{\log(p+1)}+2p+1)$
    \item $d$-map: $(d+3\floor{d/2}-3)\cdot (p+1)^2(p\ceil{\log(p+1)}+2p+1)$
    % \item $(g,d)$-map: $g4^{O(d)}$
    % \item $k$-NN: $4^{O(k^6)}$
    \item $(g,d)$-map: $\max\{2g,3\}\cdot 7d(d-3)(p+1)\cdot \binom{p+9}{9}$
  \end{compactenum}
\note{DW}{I prefer bullets here}
\end{cor}

@@

\note{DW}{This problem does not deserve a section of its own. Move to end of \cref{gk_planar_section}?}
